---
title: "Cluster Sample Size Calculations"
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---

```{r setup, echo=F}
library(tidyverse)
library(data.table)
library(CRTSize)
library(clusterPower)
```

# Plain implementations
## Function implementation from Hayes & Bennet (Int. J. Epidem. 1999;28:319-326), eqs. 1&2
```{r}
hayes_rates <- function( leff, # rate in exp. group
                         lcont, # rate in contr. group
                         alpha, beta, # significance levels
                         CV, # kappa value
                         clustersize, # average number of subjects per cluster
                         followup # follow-up time, same unit as rates
                         ) {
  zalpha <- qnorm(alpha/2, lower.tail=F) 
  zbeta <- qnorm(beta)
  #y <- (zalpha+zbeta)^2*(leff+lcont)/(leff-lcont)^2
  y <- clustersize*followup
  return( 1+(zalpha+zbeta)^2 * ((leff+lcont)/y + CV^2*(leff^2+lcont^2))/(lcont-leff)^2 )
}

hayes_rates( leff=0.01, lcont=0.005, alpha=0.05, beta=0.8, CV=0.55, clustersize=200, followup=2)
```

## Re-calculation of number of clusters per arm for an unmatched design based on Sample_size_calculations_BG-HMS.xlsx::"Rates - Unmatched"
```{r}
hayes_rates( leff=0.01036, lcont=0.0148, alpha=0.05, beta=0.8, CV=0.29, clustersize=424, followup=1)
```

## Assess sensitivity to CV (kappa) estimates by plotting number of clusters per arm against CV
```{r}
CVs <- seq(0.025, 0.5, by=0.025)

df <- data.frame(
  CV = CVs,
  clusters =sapply( CVs,
          function(x) hayes_rates( leff=0.01036, lcont=0.0148, alpha=0.05, beta=0.8, CV=x, clustersize=424, followup=1)
  )
)

ggplot( df, aes(CV, clusters) ) +
  geom_line() +
  labs( title="Influence of CV (kappa) on number of clusters per arm",
        subtitle="using Hayes & Bennet (Int. J. Epidem. 1999;28:319-326)" ) +
  geom_vline( xintercept = 0.29, linetype="dotted" ) +
  theme_classic()
```

## Alternative calculation using CRTSize (CRAN):
"Sample size estimation in cluster (group) randomized trials. Contains traditional power-based methods, empirical smoothing (Rotondi and Donner, 2009), and updated meta-analysis techniques (Rotondi and Donner, 2012)."
```{r}
n4incidence(le=0.01036, lc=0.0148, m=424, t=1, CV=0.29)
```

# Power analysis using clusterPower (CRAN)
The within-cluster variation not given. Calculate from proposed design (i.e. 35 clusters with mean=428 subjects), assuming icc==0.05.

```{r}
varw <- crtpwr.2mean(n=428, cv=0.29, d=0.0148-0.01036, icc=0.05, m=35, method = c("taylor", "weighted"),
  tol = .Machine$double.eps^0.25 )

varw
```

## power by ICC and CV (kappa)
Heatmap
```{r warning=F}
iccs <- c(seq(0.01,0.05,by=0.01), seq(0.05, 0.5, by=0.05 ), 0.75, 2:4)
CVs <- c(seq(0.025, 0.5, by=0.025), seq(1,5,by=0.5))

mat <- matrix(, nrow=length(iccs), ncol=length(CVs))
for( icc in 1:length(iccs) )
  for( cv in 1:length(CVs))
     mat[icc, cv] <- crtpwr.2mean(power=NA, 
                                  n=428, 
                                  cv=CVs[cv], 
                                  d=0.0148-0.01036, 
                                  icc=iccs[icc], 
                                  m=35, 
                                  varw=varw, 
                                  method = c("taylor", "weighted"))
colnames(mat) <- CVs
rownames(mat) <- iccs

heatmap( mat, xlab = "CV", ylab="ICC" )
```

## individual sensitivity plots
### Set up data set
```{r}
df <-
  expand.grid( ICC = c(seq(0.01,0.05,by=0.01), seq(0.05, 0.5, by=0.05 )),
               CV = c(0.1, 0.2, 0.29, seq(0.4, 3, by=0.1)),
               varw = c(0.001, seq(0.01, 1, by=0.01))) %>% 
  mutate( power = crtpwr.2mean(power=NA, 
                               n=428, 
                               cv=CV, 
                               d=0.0148-0.01036, 
                               icc=ICC, 
                               m=35, 
                               varw=varw, 
                               method = c("taylor", "weighted")) ) %>% 
  filter( !is.na(power)) %>% 
  as.data.table()

```

### Visualizations
```{r}
ggplot( df[ICC==0.05 & varw==0.001], aes(CV, power) ) +
  geom_line() +
  scale_y_continuous( limits=c(0,1))+
  labs( title="Influence of CV (kappa) on power",
        subtitle="assuming ICC=0.05 & varw=0.001") +
  theme_classic()

ggplot( df[ICC==0.05 & varw==0.01], aes(CV, power) ) +
  geom_line() +
  scale_y_continuous( limits=c(0,1))+
  labs( title="Influence of CV (kappa) on power",
        subtitle="assuming ICC=0.05 & varw=0.01") +
  theme_classic()

ggplot( df[ICC==0.5 & varw==0.001], aes(CV, power) ) +
  geom_line() +
  scale_y_continuous( limits=c(0,1))+
  labs( title="Influence of CV (kappa) on power",
        subtitle="assuming ICC=0.5 & varw=0.001") +
  theme_classic()

ggplot( df[CV==0.29 & varw==0.001], aes(ICC, power) ) +
  geom_line() +
  scale_y_continuous( limits=c(0,1))+
  labs( title="Influence of ICC on power",
        subtitle="assuming CV=0.29 & varw=0.001") +
  theme_classic()

```